### 2016-11-17 response/response.tex

\documentclass[a4paper]{article}
\usepackage{url}
\usepackage[margin=1in]{geometry}
\newcommand{\breview}{\begin{quotation}\begin{bf}\noindent}
\newcommand{\ereview}{\end{bf}\end{quotation}}
\newcommand{\reviewbullet}[1]{\breview \begin{itemize}\item #1 \end{itemize}\ereview}
\begin{document}

Dear Editor, \\[1cm]

Attached you will find the revised version of manuscript \#M106699, titled ''Structure Preserving Model Reduction of Parametric Hamiltonian Systems''.

Yours Sincerely,\\[1.0cm]
Babak Maboudi Afkham and
Jan S. Hesthaven
\\[1cm]

\section*{Reviewer 1}

\breview
1. There is a global lack of rigour in the mathematical presentation: It starts with the very first two equations. If (1) is clear, the definition of M in the equation (2) is obscure: what is the role of the time? Is there a finite time horizon? Then in equation (3) what are the v\_i? are they made of snapshots as explained later? But then what is the role of the numerical discretisation ?
\ereview

We revised major parts of the background and the introduction of the manuscript and corrected the mathematical notations for better comprehension. The authors believe that the introduction of the solution manifold, equation (2), and the reduced basis, equation (3), is essential in the understanding of model reduction techniques. The role of time was missing from the definition which is now added. However, the generation of the reduced basis, v\_i, is carefully and in much detail is discussed in chapters 3 and 4.

\breview
2. Similarly, in section 4, there are some notations used that are never defined (line 376 what do the assumptions on a and b mean?) All this lack of rigorous explanation (which contrasts with the heavy Hamiltonian formalism used in Section 3) makes the paper difficult to read and understand. 
\ereview

The explanation of the symplectic Gram-Schmidt routine (the part pointed by the reviewer) is revised for better understanding.

\breview
3. I was expecting much more convincing numerical experiments. The first stochastic example is not really understandable.

The example concerning the nonlinear Schroedinger equation is also relatively easy. If I understand correctly, there are 500 points in the parameter space and 90 in the reduced basis. Is there really a gain between computing all the trajectories and finding the reduced basis?
\ereview

We agree with the reviewer that the numerical examples presented are relatively simple. However, alternative methods are either unstable or too expensive for even these simple test cases. Therefore, a more complex numerical experiment does not contribute to a better understanding of the method.

\breview
4. I cannot find out what is plotted on Figure 1. The expectation of the solution? One realization? Then the authors talk about a gain of computational cost (which should be the goal of this new algorithm) but nothing is really shown.

What about a comparison between the different computational cost of the algorithm? 
\ereview

The purpose of Figure 1 is to illustrate how the proposed method can preserve stability with the eyeball norm, and so using just a single realization. To eliminate any confusion, we have added Figure 5 that presents the maximum error in time and parameter domain. Also the qualitative computation cost is discussed in Section 5.

\breview
A comparison with Monte Carlo algorithm is also missing. Moreover, t = 2 is not a long time for a 1D linear wave equation and the gain of symplectic algorithms is unclear. 
\ereview

The purpose of model reduction is to construct approximated solution with substantial lower cost. It is not clear to the authors, possibly due to a miss understanding, how Monte Carlo method enters this scope.

\breview
What about an illustration of the exponential convergence result given in Section 4? 
\ereview

We agree with the reviewer that a better illustration of the exponential convergence is needed. We added section 5.3 to numerically discuss this matter.

\section*{Reviewer 2}

\breview
1. The topic of projection methods for Hamiltonian systems is a widely studied topic in the context of optimal control problems where almost all the concepts discussed in this paper have already been studied, symplectic QR, symplectic SVD, symplectic Krylov and Lanc- zos techniques, model reduction for Hamiltonian systems (Riccati equations) etc. All these techniques are reinvented in this paper without reference to the previous work. This is unac- ceptable and in some sense even unethical, since any google search would have immediately revealed these papers by e.g. Benner, Breiten, Bunse-Gerstner, Freund, Kressner, Mehrmann, Poloni, Saak, Schoeder, Simoncini, Watkins etc. Many of these are also dealing with model reduction and particularly adress the preservation of Hamiltonian and symplectic structure in projection methods.
\ereview

Most of the topics and references introduced above deal with eigenvalue problems of Hamiltonian, skew-Hamiltonian and symplectic matrices. The main focus of this manuscript however, is the model reduction of Hamiltonian systems specifically the greedy approach to structure preserving model reduction. This, although strongly related, is not reflected in the references suggested by the reviewer. Nonetheless we agree that including these topics in the paper is essential as they comprise a major part of the development of the field in optimal control. Therefore, most of these references are included in the revision of the paper.

\breview
2. There are also some mathematical questions that need to be addressed. It is well-known that symplectic bases are not norm bounded, so special precautions have to be taken to guarantee that they do not become unbounded. This can be done by the isotropic Arnoldi/Gram Schmidt and needs to be done here. Unfortunately this procedure requires the existence of Lagrange subspaces which is not the case here.

I do not see the reflection of this unboundedness in the convergence proof, I guess it is reflected in the constant, but this should be analyzed. 
\ereview
The symplectic greedy method suggested in the paper, produces an orthosymplectic (symplectic and orthonormal) basis by construction. The orthonormality of the basis guarantees a numerically stable basis. Although this was briefly discussed in the paper, there was clearly a need for a detailed discussion. To this end, it is shown in the revision of the paper, as pointed by the reviewer, that the greedy method constructs a Lagrangian subspace in each iteration and then expands it to an orthosymplectic basis. We emphasised on this in the revision of the paper in form of Theorem 21.

\breview
Another mathematical problem is the stability of the projected system, this is to my opinion not guaranteed and needs to be proved, e.g. it needs to be guaranteed that there does not exists a zero Lyapunov exponent associated with crossing trajectories and with mixed sign/Krein characteristic which puts the Hamiltonian system on the boundary of the stable region and small perturbations will make it unstable. Here structure preservation helps but does not solve the problem.
\ereview

It was briefly discussed in the paper that stable critical points of a Hamiltonian system remains stable in the reduced system. This matter is discussed in more detail in the revision of the paper. This together with the numerical experiments in the paper, although do not guarantee general stability, but are strong indication of the stability of the reduced system.

\end{document}

### 2016-11-17 greedy_hamiltonian_revised/chapters/3.ham.tex
all done, response to the referees also added

\section{Hamiltonian Systems and Symplectic Geometry} \label{chap:Hasy:1}
Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate 2-form on $\edit \mathcal M$. The pair $\edit (\mathcal M,\Omega)$ is called a \emph{symplectic manifold} {\edit \cite{Marsden:1999ck}}.

@@

{\edit
\begin{theorem} \label{theorem:Hasy:1.1}
If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$ \cite{Marsden:1999ck}. 
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.2}
If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional \cite{Marsden:1999ck}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.3}
(The Symplectic Gram-Schmidt) If $(V,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $V$ such that
\begin{equation} \label{eq:Hasy:4}
\begin{aligned}
	& \Omega(e_i,e_j) = 0 = \Omega(f_i,f_j), \quad & i\neq j,\\
	& \Omega(e_i,f_j) = \delta_{ij}, & i\leq i,j \leq n.
\end{aligned}
\end{equation}
where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose the basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
\begin{equation} \label{eq:Hasy:5}
	\Omega(v_1,v_2) = v_1^T \mathbb J_{2n} v_2, \qquad v_1,v_2\in \mathbb R^n,
\end{equation}
with $\mathbb J_{2n}$ being the symplectic matrix, defined as
\begin{equation} \label{eq:Hasy:6}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix}.
\end{equation}
Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively \cite{de2006symplectic}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.4}
The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as \cite{Marsden:1999ck}
\begin{equation}
	\langle v,u \rangle = \Omega(\mathbb J_{2n}v,u),\quad \forall u,v \in \mathbb R^{2n}.
\end{equation}
\end{theorem}
}

{\edit 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $P$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general \cite{de2006symplectic}. 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$, where $dim$ denotes the dimension of the subspace \cite{abraham1978foundations}.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}
Let $E\subset V$ be a Lagrangian vector space. Then there is an orthosymplectic basis for $\mathbb R^{2n}$ \cite{da2003introduction}.
\end{theorem}
\begin{proof}
Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed.  Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define the new set of vectors $f_1 = -\mathbb Je_1$, $f_2 = -\mathbb Je_2$, $\dots$, $f_n= -\mathbb Je_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n}^T \mathbb J_{2n} e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = -e_i \mathbb J^T_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes \cite{Marsden:1999ck}
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

@@

{\edit It natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}. Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a wrong qualitative behavior in the solution \cite{Hairer:1250576}. Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure yielding stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
\begin{equation} \label{eq:Hasy:13}
\begin{aligned}
	q_{n+1/2} &= q_n + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_n), \\
	p_{n+1} &= p_n - \frac{\Delta t}{2} \left( \nabla_qH(q_{n+1/2},p_n) + \nabla_qH(q_{n+1/2},p_{n+1}) \right),\\
	q_{n+1} &= q_{n+1/2} + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_{n+1}),
\end{aligned}
\end{equation}
and
\begin{equation} \label{eq:Hasy:14}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
For a general Hamiltonian system, Str\"omer-Verlet scheme is implicit. However, for separable Hamiltonians, i.e. $H(q,p) = K(p) + U(q)$, this {\edit scheme becomes} explicit. We refer the reader to \cite{Hairer:1250576} for more information about the construction and applications of symplectic and geometric numerical integrators.

>>>>

\section{Hamiltonian Systems and Symplectic Geometry} \label{chap:Hasy:1}
Let $\edit \mathcal M$ be a manifold and $\edit \Omega:\mathcal M \times \mathcal M \to \mathbb R$ be a closed, nondegenerate {\edit and skew-symmetric} 2-form on $\edit \mathcal M$. The pair $\edit (\mathcal M,\Omega)$ is called a \emph{symplectic manifold} {\edit \cite{Marsden:1999ck}}.

@@

{\edit
\begin{theorem} \label{theorem:Hasy:1.1}
If $V$ is a vector space then $\Omega$ is a constant form, that is $\Omega_z$ is independent of $z\in V$ \cite{Marsden:1999ck}. 
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.2}
If $(V,\Omega)$ is a finite-dimensional symplectic manifold then $V$ is even dimensional \cite{Marsden:1999ck}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.3}
(The Symplectic Gram-Schmidt) If $(V,\Omega)$ is a $2n$-dimensional symplectic vector space, then there is a basis $e_1,\dots e_n,f_1, \dots , f_n$ of $V$ such that
\begin{equation} \label{eq:Hasy:4}
\begin{aligned}
	& \Omega(e_i,e_j) = 0 = \Omega(f_i,f_j), \quad & i\neq j,\\
	& \Omega(e_i,f_j) = \delta_{ij}, & i\leq i,j \leq n.
\end{aligned}
\end{equation}
where $\delta$ is the Kronecker's delta function. Moreover if $V = \mathbb{R}^{2n}$ we can choose basis vectors $\{e_i,f_i\}_{i=1}^n$ such that
\begin{equation} \label{eq:Hasy:5}
	\Omega(v_1,v_2) = v_1^T \mathbb J_{2n} v_2, \qquad v_1,v_2\in \mathbb R^n,
\end{equation}
with $\mathbb J_{2n}$ being the symplectic matrix, defined as
\begin{equation} \label{eq:Hasy:6}
	\mathbb{J}_{2n} = 
	\begin{pmatrix}
		0_n & I_n \\
		-I_n & 0_n
	\end{pmatrix}.
\end{equation}
Here $I_n$ and $0_n$ is the identity matrix and the zero square matrix of size $n$, respectively \cite{de2006symplectic}.
\end{theorem}
\begin{theorem} \label{theorem:Hasy:1.4}
The classical inner product $\langle \cdot,\cdot \rangle:\mathbb R^{2n}\times \mathbb R^{2n}\to \mathbb R$ can be written in terms of the 2-form as \cite{Marsden:1999ck}
\begin{equation}
	\langle v,u \rangle = \Omega(\mathbb J_{2n}v,u),\quad \forall u,v \in \mathbb R^{2n}.
\end{equation}
\end{theorem}
}

{\edit 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space and $E\subset V$ is a subspace. Then the symplectic complement of $E$ inside $P$ is defined as
\[
	E^{\perp} := \{ v\in V |\ \Omega(v,e) = 0,\ \forall e\in E \}
\]
\end{definition}
Note that $E \cap E^{\perp}$ is not empty in general \cite{de2006symplectic}. 
\begin{definition}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. A subspace $E\subset V$ is called a Lagrangian subspace inside $V$ if $E = E^\perp$.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.5}
Suppose $(V,\Omega)$ is a finite dimensional symplectic vector space. If $E\subset V$ is a Lagrangian subspace then $dim(E)=\frac 1 2dim(V)$, where $dim$ denotes the dimension of the subspace \cite{abraham1978foundations}.
\end{theorem}
\begin{definition}
A basis of $(V,\Omega)$ is called orthosymplectic if it is both a symplectic basis and an orthogonal basis with respect to the classical scalar product.
\end{definition}
\begin{theorem} \label{theorem:Hasy:1.6}
Suppose $(V,\Omega)$ is a $2n$ dimensional symplectic vector space and $E\subset V$ is a Lagrangian subspace. Then there is an orthosymplectic basis for $V$ \cite{da2003introduction}.
\end{theorem}
\begin{proof}
Starting from a Lagrangain subspace in $E \subset V$ an orthosymplectic basis can be easily constructed. By Theorem \ref{theorem:Hasy:1.5} $E$ is $n$ dimensional. Suppose that $\{ e'_1,\dots, e'_n \}$ is a basis for $E$, using the classical Gram-Schmidt orthogonalization process we can construct an orthonormal basis $\{ e_1,\dots,e_n \}$. Define the new set of vectors $f_1 = \mathbb J_{2n}^Te_1$, $f_2 =\mathbb J_{2n}^T e_2$, $\dots$, $f_n= \mathbb J_{2n}^Te_n$. We have
\begin{equation}
	\langle f_i, f_j \rangle = e_i^T \mathbb J_{2n} {\mathbb J_{2n}}^T e_i = \delta_{ij}, \quad \langle f_i, e_j \rangle = e_i^T \mathbb J_{2n} e_j = 0, \quad i,j=1,\dots,n,
\end{equation}
where we used the fact that $\mathbb J_{2n} {\mathbb J_{2n}}^T = I_{2n}$ in the first identity and the second identity is due to the fact that the basis $\{ e_1,\dots,e_n \}$ forms a Lagrangian subspace. This shows that the set $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ forms an orthonormal basis. Also, it can be easily verified that this is a symplectic basis. Thus $\{ e_1,\dots,e_n \}\cup \{ f_1,\dots,f_n \}$ is an orthosymplectic basis.
\end{proof}
\begin{theorem} \label{theorem:Hasy:1.7}
On a finite-dimensional symplectic vector space the relationship (\ref{eq:Hasy:1}) becomes \cite{Marsden:1999ck}
\begin{equation} \label{eq:Hasy:7}
\left\{
\begin{aligned}
	&\dot {\mathbf z} = \mathbb{J}_{2n} \nabla_{\mathbf z} H(\mathbf z), \\
	& \mathbf z(0) = \mathbf z_0.
\end{aligned}
\right.
\end{equation}
or, by introducing the canonical coordinates $\mathbf z = (\mathbf q^T, \mathbf p^T)^T$,
\begin{equation} \label{eq:Hasy:8}
\left\{
\begin{aligned}
	&\dot {\mathbf q} = \nabla_{\mathbf p} H(\mathbf q,\mathbf p),\\
	&\dot {\mathbf p} = - \nabla_{\mathbf q} H(\mathbf q,\mathbf p).
\end{aligned}
\right.
\end{equation}
\end{theorem}
}

@@

{\edit It is natural to expect a numerical integrator that solves (\ref{eq:Hasy:7}) to also satisfy the conservation law in Theorem \ref{theorem:Hasy:1}. Common numerical integrators e.g., Runge-Kutta methods, do not generally preserve the Hamiltonian which results in a wrong qualitative behavior in the solution \cite{Hairer:1250576}. Symplectic integrators are a class of numerical integrators for Hamiltonian systems that preserve the symplectic structure yielding stability in long-time integration.} The Str\"omer-Verlet time stepping scheme is an example of symplectic integrators and is given by
\begin{equation} \label{eq:Hasy:13}
\begin{aligned}
	q_{n+1/2} &= q_n + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_n), \\
	p_{n+1} &= p_n - \frac{\Delta t}{2} \left( \nabla_qH(q_{n+1/2},p_n) + \nabla_qH(q_{n+1/2},p_{n+1}) \right),\\
	q_{n+1} &= q_{n+1/2} + \frac{\Delta t}{2} \nabla_pH(q_{n+1/2},p_{n+1}),
\end{aligned}
\end{equation}
and
\begin{equation} \label{eq:Hasy:14}
\begin{aligned}
	p_{n+1/2} &= p_n - \frac{\Delta t}{2} \nabla_qH(q_{n},p_{n+1/2}), \\
	q_{n+1} &= q_n + \frac{\Delta t}{2} \left( \nabla_pH(q_{n},p_{n+1/2}) + \nabla_pH(q_{n+1},p_{n+1/2}) \right),\\
	p_{n+1} &= p_{n+1/2} - \frac{\Delta t}{2} \nabla_qH(q_{n+1},p_{n+1/2}).
\end{aligned}
\end{equation}
For a general Hamiltonian system, Str\"omer-Verlet scheme is implicit. However, for separable Hamiltonians, i.e. $H(q,p) = K(p) + U(q)$, this {\edit scheme becomes} explicit. We refer the reader to \cite{Hairer:1250576} for more information about the construction and applications of symplectic and geometric numerical integrators.
